---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Following code loads data from **activity.csv** file and saves it in *activity_data* variable.

```{r read_data}
# read data from activity.csv file
activity_data <- read.csv("activity.csv")
# convert date from Factor to Date format
activity_data$date <- as.Date(as.character(activity_data$date), format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Following code is for computing number of steps taken per day.

```{r steps_per_day}
# calculate total number of steps taken per day
steps_per_day <- tapply(activity_data$steps, factor(activity_data$date), FUN = mean)
# plot histogram of steps taken per day
hist(steps_per_day)
# compute mean of steps taken per day
mean_steps_per_day <- mean(steps_per_day, na.rm = TRUE)
# compute median of steps taken per day
median_steps_per_day <- median(steps_per_day, na.rm = TRUE)
```

The mean of number of steps taken per day is: `r mean_steps_per_day`.  
The median of number of steps taken per day is: `r median_steps_per_day`.

## What is the average daily activity pattern?

Following code creates a time-series plot of average number of steps taken per day.

```{r timeseries_of_average_across_all_days}
# calculate average steps per interval over all days
steps_averaged_across_all_days <- tapply(activity_data$steps, factor(activity_data$interval), function(x) mean(x, na.rm = TRUE))
# plot time series data of average_steps_per_day vs interval
plot(unique(activity_data$interval), steps_averaged_across_all_days, type = "l")
```


## Imputing missing values

Following code calculates the number of rows in the dataset containing missing values.

```{r count_rows_with_missing_values}
# calculate total number of missing values (NAs) in the dataset
no_of_rows_with_missing_values <- length(complete.cases(activity_data)[complete.cases(activity_data) == FALSE])
```

The total number of missing values in the dataset (i.e. the total number of rows with NAs) : `r no_of_rows_with_missing_values`.  

Lets replace all the missing values in steps column with the mean of overall steps values.  
Create a new dataset called updated_activity_data from activity_data with NAs replaced.

```{r replace_NAs}
# replace NAs in steps column with mean of steps
mean_of_steps_column <- mean(activity_data$steps, na.rm = TRUE)
updated_activity_data <- activity_data
updated_activity_data$steps[is.na(updated_activity_data$steps)] <- mean_of_steps_column
```

Lets now create a histogram for steps_per_day in updated dataset.
```{r updated_steps_per_day}
# calculate total number of steps taken per day for updated dataset
updated_steps_per_day <- tapply(updated_activity_data$steps, factor(updated_activity_data$date), FUN = mean)
# plot histogram of steps taken per day for updated dataset
hist(updated_steps_per_day)
# compute mean of steps taken per day in updated dataset
updated_mean_steps_per_day <- mean(updated_steps_per_day)
# compute median of steps taken per day in updated dataset
updated_median_steps_per_day <- median(updated_steps_per_day)
```

The mean of number of steps taken per day in updated dataset is: `r updated_mean_steps_per_day`.  
The median of number of steps taken per day in updated dataset is: `r updated_median_steps_per_day`.

These values do not differ from the same from original dataset as we have replaced missing values for steps column with the **mean of entire steps column**.  

## Are there differences in activity patterns between weekdays and weekends?

